---
title: "WHO Measles Investigation"
format: 
  dashboard:
    orientation: rows
server: shiny
embed-resources: true
logo: logo.png
logo-alt: "World Health Organization logo with globe and medical snake."
theme: flatly
nav-buttons:
      - icon: github
        href: https://github.com/nanamiikii/STAT541_Project2
        aria-label: GitHub
---

```{r}
#| label: load-packages
#| echo: false
#| output: false
#| context: setup

# quality of life
library(tidyverse)

# tables 
library(gt)

# json files
library(tidyjson)
library(jsonlite)

# webscraping/api
library(httr)

# modeling
library(nlme)

# visualizations
library(rnaturalearthdata)
library(rnaturalearth)
library(tidyverse)
library(ggtext)
library(scales)
library(fmsb)
library(ggalluvial)
library(ggrepel)
library(htmlwidgets)
library(leaflet)
library(gganimate)
library(gifski)
library(shiny)
```

```{r}
#| label: read in data from GitHub
#| output: false
#| eval: false

cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
```

```{r}
#| label: creating function with iteration for summary statistics 

########## HELPER FUNCTIONS ##########
######################################

# using the same summ_stats as practice activity 2
summ_stats <- function(x) {
    
    # if x contains numeric values
      # return a tibble with the mean, med, sd, IQR, and n observed
    if (is.numeric(x)) {
        out_df <- tibble(
            mean = mean(x, na.rm = TRUE), 
            median = median(x, na.rm = TRUE), 
            stdev = sd(x, na.rm = TRUE), 
            iqr = IQR(x, na.rm = TRUE),
            n = sum(!is.na(x)) # length won't work bc of NAs
        )
    # if x contains categorical, return tibble with # of levels in variable 
    } else if (is.factor(x) || is.character(x)) {
        out_df <- tibble(
            n_levels = n_distinct(x)
        )
    } else {
        stop("Inputted values are not numeric or categorical.")
    }

    #return the data frame.
    out_df
}

summarize_stats <- function(df, col = measles_lab_confirmed, mode = region) {
  ######### PARAMETERS ##########
  # input: df -- the dataframe of interest 
  # col: column of interest to summarize by: 
  #   measles_suspect, 
  #   measles_clinical, 
  #   measles_epi_linked, 
  #   measles_lab_confirmed [DEFAULT], 
  #   measles_total
  # mode: aggregate by options: 
  #   region [DEFAULT], 
  #   country, 
  #   iso3, 
  #   year 
  ###############################
  
  df |>
    select({{mode}}, {{col}}) |>
    group_by({{mode}}) |>
    #group_map removed the group column, so to retain that, group_modify.
    # See second example: https://dplyr.tidyverse.org/reference/group_map.html
    group_modify( ~ summ_stats(.x |>
                              pull({{col}}))) |>
    bind_rows()
}
```

```{r}
#| title: api calling
#| eval: false

# API call
data <- GET("https://www.imf.org/external/datamapper/api/v1/NGDPDPC")

# creating function to make tibbles for each country from json data
country_values <- function(country){
  
  tibble(json = list(parse_json(data))) |> 
    unnest_wider(json) |> 
    unnest_longer(values) |> 
    unnest_wider(values) |>
    slice_head() |> 
    select({{country}}) |> 
    unnest_longer({{country}}) |> 
    rename(year = str_glue("{country}_id")) |> 
    pivot_longer(cols = {{country}},
                 names_to = "iso3",
                 values_to = "GDP_per_capita"
                 ) |> 
    mutate(year = as.numeric(year))
  
}

# storing countries available in both the cases_year dataset and json data
ids <- cases_year$iso3 |>
  as.factor() |>
  levels()

ids <- ids[ids %in% (
  tibble(json = list(parse_json(data))) |> 
    unnest_wider(json) |> 
    unnest_longer(values) |> 
    unnest_wider(values) |>
    slice_head() |> 
    names()
  )]

# creating dataset from json data for each country and merging to one
data <- map(.x = ids,
            .f = country_values
            ) |> 
  bind_rows()

# join to cases_year dataset to include all data
dataset <- cases_year |> 
  full_join(data) |> 
  filter(year >= 2012) #|> 
  #mutate(region_info = fct_recode(
    #region,
    #"Africa" = "AFRO",
    #"Americas" = "AMRO",
    #"Eastern Mediterranean" = "EMRO",
    #"Europe" = "EURO",
    #"South East Asia" = "SEARO",
    #"Western Pacific" = "WPRO"
    #)
    #)
```

```{r}
#| label: API call 2
#| eval: false
data <- GET("https://ghoapi.azureedge.net/api/MCV2")

data <- tibble(json = list(parse_json(data))) |> 
  unnest_wider(json) |> 
  unnest_wider(value, names_sep = "_") |>
  select(-"@odata.context")
```

```{r}
#| label: creating function to extract data from json data
#| eval: false
row <- function(v){
  data |> 
    select(v) |> 
    unnest_wider(v) |> 
    select(ParentLocationCode,
           #ParentLocation,
           SpatialDim,
           TimeDim,
           #Value,
           NumericValue
           )
}
```

```{r}
#| label: mapping data 
#| eval: false
data <- map(.x = data |> names(),  
            .f = row
            ) |> 
  bind_rows() |> 
  rename(
    "region" = ParentLocationCode,
    #"region_info" = ParentLocation,
    "iso3" = SpatialDim,
    "year" = TimeDim,
    #value = Value,
    "mcv2" = NumericValue
  )
```

```{r}
#| label: creating big metadata
#| eval: false

metadata <- dataset |> 
  
  # join to cases_year dataset
  full_join(data |> select(-region),
            by = join_by(iso3, year)
            ) |> 
  filter(year >= 2012) |> 
  mutate(mcv2 = mcv2 / 100) |> 
  
  # recoding region to be more informative
  mutate(
    region_s = region,
    region = fct_recode(
      region,
      "Africa" = "AFRO",
      "Americas" = "AMRO",
      "Eastern Mediterranean" = "EMRO",
      "Europe" = "EURO",
      "South East Asia" = "SEARO",
      "Western Pacific" = "WPRO"
      )
    ) |> 
  
  mutate(p_lab_total = measles_lab_confirmed / measles_total)

```

```{r}
#| label: loading in data
#| context: setup

load(here::here("data", "metadata.RData"))
```

# Measles and Region Socioeconomic Indicators

## Row 1 - research question {height="15%"}

:::{.valuebox color="#a2c7ea"}

[{{< fa clipboard-question >}} **Are there associations between measles and region socioeconomic indicators?**]{style="font-size:24pt"}

<!-- For this project, our aim is to see what socioeconomic factors, such as those relating to income, seem to contribute to the number of measles for the different world regions as defined by the World Health Organization (WHO). -->

:::

## Row 2 - plots and tables {height="85%"}

### Column 1 - plots {.tabset width="65%"}

::: {.card title="2024 MCV2 Rates"}

```{r}
#| label: leaflet plot
#| title: 2024 Measles Vaccination Rates Around the World
#| context: setup
#| fig-alt: An interactive world map displaying 2024 measles vaccination rates on a red color scale. Hovering over each country provides the Many countries in North America, Europe, and parts of Asia appear lighter, suggesting higher vaccination coverage, while several countries in central and sub-Saharan Africa are darker, suggesting lower coverage. Some countries are gray, representing unknown vaccination rates.
metadata_2024 <- metadata |>
  filter(year == 2024)

selectInput("region_select",
            "Select a region:",
            choices = c("World",
                        sort(unique(as.character(metadata_2024$region)))),
            selected = "World")

leafletOutput("map")
```

An interactive world map displaying 2024 measles vaccination rates on a red color scale. Hovering over each country provides the country name and MCV2 vaccination rate. Many countries in *North America*, *Europe*, and parts of *Asia* appear lighter, suggesting higher vaccination coverage, while several countries in *central and sub-Saharan Africa* are darker, suggesting lower coverage.

:::

```{r}
#| context: server

map_data <- ne_countries(scale = "medium", returnclass = "sf") |>
  left_join(metadata_2024, join_by(iso_a3_eh == iso3))

pal <- colorNumeric(palette = c("#e74c3c", "#ecf0f1"),
                    domain = c(0, 1),
                    na.color = "darkgray")

llz <- reactive({ # list of long, lat, zoom for setView()
  switch(input$region_select,
    "World" = c(0, 25, 1),
    "Africa" = c(20, 0, 3),
    "Americas" = c(-80, 10, 3),
    "Eastern Mediterranean" = c(45, 25, 4),
    "Europe" = c(10, 50, 4),
    "South East Asia" = c(100, 15, 4),
    "Western Pacific" = c(130, 0, 3)
  )
})

output$map <- renderLeaflet({
  coords <- llz()
  map_data |>
    leaflet() |>
    setView(lng = coords[1], lat = coords[2], zoom = coords[3]) |>
    onRender("function(el, x) {el.style.backgroundColor = '#3498db';}") |>
    addPolygons(fillColor = ~pal(mcv2),
                fillOpacity = 1,
                color = "white",
                weight = 1,
                opacity = 1,
                label = ~paste0(name, ": ",
                                ifelse(is.na(mcv2),
                                       "No Data",
                                       scales::percent(mcv2))))
})
```


::: {.card title="Measles and MCV2 By Income"}

```{r}
#| label: time series plot
#| title: measles and Vaccination By Income

metadata |>
  select(region, country, year, measles_incidence_rate_per_1000000_total_population, total_population, mcv2, GDP_per_capita) |>
  filter(year != 2025 & !is.na(region) & !is.na(GDP_per_capita)) |>
  mutate(gdp_group = factor(
          case_when(
            GDP_per_capita <= 1145 ~ "Low", 
            GDP_per_capita <= 14005 ~ "Middle", 
            GDP_per_capita > 14005 ~ "High"
            ), 
          levels = c("Low", "Middle", "High")
          )
        ) |>
  group_by(gdp_group, year) |>
  summarize(Vaccination = median(mcv2 * 100, na.rm = TRUE), 
            Incidence = median(measles_incidence_rate_per_1000000_total_population, na.rm = TRUE), 
            .groups = "drop") |>
  pivot_longer(
    cols = c(Incidence, Vaccination), 
    names_to = "type", 
    values_to = "rate"
  ) |>
  ggplot(aes(x = year, y = rate, color = type)) + 
  geom_line() + 
  scale_color_manual(values = c("Incidence" = "#e74c3c", "Vaccination" = "#3498db")) + 
  facet_wrap(~ gdp_group, axes = "all") + 
  labs(
    title = "Comparison of Median <span style='color:#e74c3c;'>Incidence Rate per Million</span> and <span style='color:#3498db;'>MCV2 Vaccination Percentage</span>",
    subtitle = "Countries grouped by income level defined as by World Bank",
    x = "Year",
    y = "",
    alt = "A figure that displays line charts with median measles incidence and MCV2 vaccination rates per million people from 2012 to 2024, faceted by income classification. Vaccination rate are generally higher as the income classification group rises, which indicates generally that richer countries are more vaccinated. Low income countries have a more direct relationship between these two variables.",
    color = "Type of Rate") +
  scale_x_continuous(breaks = seq(2012, 2024, by = 4), 
                    ) + 
  theme_minimal() + 
  theme(
    plot.title = element_markdown(size = 12), 
    legend.position = "none"
    )

```

Vaccination rate are generally *higher* as the income classification group rises, which indicates generally that richer countries are more vaccinated. Low income countries have a more direct relationship between these two variables.

:::

### Column 2 - tables {.tabset width="35%"}

```{r}
#| label: table 1
#| title: GDP Per Capita Summary

gdp_groups <- metadata |>
  filter(year == 2024, !is.na(region), !is.na(GDP_per_capita)) |>
  group_by(region) |>
  summarize(mean_gdp = mean(GDP_per_capita, na.rm = TRUE), .groups = "drop") |>
  mutate(
    gdp_group = factor(
      case_when(
        mean_gdp <= 1145 ~ "low",
        mean_gdp <= 14005 ~ "middle",
        TRUE ~ "high"
      ),
      levels = c("low", "middle", "high")
    )
  )

summarize_stats(metadata |> filter(year == 2024), col = GDP_per_capita) |>
  filter(!is.na(region)) |>
  left_join(gdp_groups, by = "region") |>
  arrange(desc(mean)) |> 
  ungroup() |>
  gt() |> 
  fmt_currency(columns = -c(region, n), decimals = 2) |> 
  cols_align(align = "left", columns = region) |>
  cols_label("region" = "Region",
              "mean" = "Mean",
              "median" = "Median",
              "stdev" = "Standard Deviation",
              "iqr" = "IQR",
              "n" = "Sample Size",
              ) |> 
  tab_style(
    style = cell_fill(color = "#e74c3c50"),
    locations = cells_body(rows = gdp_group == "low")
  ) |>
  tab_style(
    style = cell_fill(color = "#f39c1250"),
    locations = cells_body(rows = gdp_group == "middle")
  ) |>
  tab_style(
    style = cell_fill(color = "#3498db50"),
    locations = cells_body(rows = gdp_group == "high")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
    ) |> 
  tab_style(
    style = cell_text(style = "italic", decorate = "underline"),
    locations = cells_row_groups()
    ) |>
  cols_hide(columns = c(gdp_group, mean_gdp)) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
    ) |>
  tab_header(
    title = md("Summary Statistics for GDP Per Capita (USD) by Region in 2024"), 
    subtitle = md("GDP classification based on mean of region: 
      <span style='color:#e74c3c;'>Low</span> | <span style='color:#f39c12;'>Middle</span> | <span style='color:#3498db;'>High</span>")
  )

```

```{r}
#| label: table 2
#| title: GDP per Capita and MCV2 Rate

metadata |>
  #filter_out(is.na(region)) |>
  filter(!(is.na(region))) |> 
  group_by(region) |>
  summarize(mean_GDP = mean(GDP_per_capita, na.rm = TRUE),
            mean_mcv2 = mean(mcv2, na.rm = TRUE)) |>
  arrange(mean_GDP) |>
  gt() |>
  data_color(columns = mean_GDP, method = "numeric",
             fn = col_numeric(palette = c("#ecf0f1", "#3498db"),
                              domain = NULL)) |>
  data_color(columns = mean_mcv2, method = "numeric",
             fn = col_numeric(palette = c("#ecf0f1", "#29c1a3"),
                              domain = NULL)) |>
  fmt_currency(columns = mean_GDP, currency = "USD") |>
  fmt_percent(columns = mean_mcv2, decimals = 1) |>
  cols_align(align = "left", columns = region) |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  tab_header(title = md("Average GDP per Capita and MCV2 Vaccine Rate by Region")) |>
  tab_caption(caption = md("Side-by-side comparison of mean GDP and MCV2 Percentage")) |>
  cols_label(region = "Region", mean_GDP = "Mean GDP (USD)", mean_mcv2 = "Mean MCV2 Coverage")
```

# Mongolia

## Row 1 - why mongolia {height="28%"}

:::{.valuebox color="#a2c7ea"}

[{{< fa circle-question >}} **Why Mongolia?**]{style="font-size:16pt"}

We are interested in looking at Mongolia as from the table below, we see that overall, they have the highest mean number of measles cases per 1 million people, yet based on the map on the main page, they appear to have a very high vaccination rate in 2024 (95%). Knowing that vaccines tend to reduce the transmission and transfection of measles, we wanted to investigate this country further to gather more insight on possible reasons for this inconsistency in expected trend and also to see the association between number of measles and infant mortality rate.

[{{< fa circle-exclamation >}} **Key Takeaway**]{style="font-size:16pt"}

According to the plot, Mongolia’s mortality for infants and children under 5 years old appears to be decreasing over time, while the country does not demonstrate a change in the percentage of measles cases with the exception of a spike in 2025 and 2026. Therefore, there does not seem to be an association between this spike and the infant mortality rate for the country.
:::

## Row 2 - table and animation {height="72%"}

### Column 1 - table {width="45%"}

```{r}
#| title: Measles Cases Across the Regions of the World

metadata |> 
  filter(!(is.na(region)),
         !(is.na(country))
         ) |> 
  
  # used to assign each country to the appropriate region when merging
  select(region, country) |> 
  unique() |>
  arrange(country) |> 
  
  full_join(summarize_stats(metadata |> filter(!(is.na(region)), !(is.na(country))),
                            measles_incidence_rate_per_1000000_total_population,
                            country)) |> 
  group_by(region) |> 
  slice_max(mean, n = 5) |> 
  arrange(-mean) |> 
  gt() |> 
  fmt_number(columns = -c(country, region, n), 
             decimals = 2) |> 
  cols_label("country" = "Country",
             "mean" = "Mean",
             "median" = "Median",
             "stdev" = "Standard Deviation",
             "iqr" = "IQR",
             "n" = "Sample Size",
             ) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
    ) |> 
  tab_style(
    style = cell_text(style = "italic", decorate = "underline"),
    locations = cells_row_groups()
    ) |> 
  tab_header(
    title = md("Summary Statistics for Total Number of Measles Cases for Each
               1 Million Individuals")
  ) |> 
  data_color(
    columns = "mean",
    method = "numeric",
    palette = c("#CCEEDF", "#BAE8D4", "#A8E2CA",
                "#96DCBF", "#83D6B5", "#6ED0AB",
                "#57CAA1", "#3BC397", "#00BD8D"
                )
    )
```

### Column 2 - animated plot

```{r}
#| title: Percentage of Measles Cases and Mortality Over Time

metadata |> 
  right_join(
    read_csv(here::here("data", "2322814_ALL_LATEST.csv")) |>
      filter(
        GEO_NAME_SHORT == "Mongolia", 
        DIM_TIME >=  2012, 
        DIM_TIME <= 2024, 
        DIM_POP_WEALTH_QUINTILE == "TOTAL",
        DIM_SEX == "TOTAL" # possibly remove later, there is data for SAAB
        ) |>
      arrange(desc(DIM_TIME)),
    by = c("country" = "GEO_NAME_SHORT", "year" = "DIM_TIME")
    ) |> 
  select(region, country, year, measles_total, total_population,
         #measles_incidence_rate_per_1000000_total_population,
         mcv2,
         RATE_PER_1000_N
         ) |> 
  filter(country == "Mongolia",
         year < 2025
         ) |> 
  rename(#"measles_r" = "measles_incidence_rate_per_1000000_total_population",
         "mortality_r" = "RATE_PER_1000_N"
         ) |> 
  mutate(measles = measles_total / total_population,
         mortality = mortality_r / 1000
         ) |> 
  pivot_longer(
    cols = c(measles,
             #mcv2,
             mortality),
    names_to = "type",
    values_to = "rate"
  ) |> 
  ggplot(aes(x = year,
             y = rate,
             color = type |> fct_relevel(
               c(
               #"mcv2",
               "mortality",
               "measles"
               )
               ) |> 
               fct_recode(
                 #"Mortality (deaths below age 5)" = "mortality",
                 "Under-5 Mortality Rate" = "mortality",
                 "Measles Rate" = "measles")
             )
         ) + 
  geom_point() +
  geom_line() +
  transition_reveal(year) + 
  #transition_time(year) + 
  labs(
    x = "Year",
    y = "",
    subtitle = "Year: {round(frame_along, 0)}",
    color = "",
    #title = "Number of measles cases and number of mortalities as percentages of Mongolia population"
    title = ""
    ) + 
  #scale_y_continuous(limits = c(0, 1)) + 
  scale_color_manual(values = c("#001E4D", "#FF9807", "#B60BE8")) + 
  scale_y_continuous(labels = label_percent()) + 
  theme_minimal()
```
